\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{hyperref}
\usepackage{subfig}
\usepackage{multicol}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{semantic}
\usepackage{stmaryrd}

\newcommand\Iota{\textsc{Iota}}
\newcommand\Menhir{\textsc{Menhir}}
\newcommand\Odoc{\textsc{Odoc}}
\newcommand\Ocaml{\textsc{OCaml}}
\newcommand\Dune{\textsc{Dune}}
\newcommand\Kappa{\textsc{Kappa}}

\usepackage{minted}
\newcommand{\ocaml}{\mintinline{ocaml}}

\usepackage{syntax}

\setcounter{tocdepth}{1}

\author{Alexandre Moine}
\title{Iota - Manual}

\begin{document}

\maketitle

\abstract{\Iota{} is well-described language with a strong type system that can be compiled to Husson's logic \cite{husson}. The compiler comes with several static analysis and optimizations. \Iota{} also provides programming facilities through a powerful inference, and the possibility to extend and constrain the type system.}

\tableofcontents

\section{Introduction}

\subsection{Molecular biology}
Molecular biology mainly studies protein-protein interactions. Such interactions are complex to understand, but even the knowledge about it is hard to represent. Here is an example borrowed from Husson's\cite{husson} thesis of a typical sentence from a molecular biology paper:

\begin{quote}
  The activation of Raf-1 by activated Src requires phosphorylation of Raf-1 on Y340 and/or
  Y341 [...]. Tyrosine phosphorylation and activation of Raf-1 have been shown to be co-incident.
  However, others have been unable to detect phosphotyrosine in active Raf-1.\cite{masson}
\end{quote}

The use of a semi-formal English makes it hard for this knowledge to be analysed and shared.
Moreover, the field of molecular biology is seeing an explosion of activity: there is so much data that human capabilities are outmatched. Taking Krivine's example\cite{siglog}, there was more than 500 paper published on the rather specific subject of ``Mitogen Activated Protein Kinases'' in the single year of 2008.\\

The combination of a high amount of data with a poor representation has led to the need and development of formal methods in Biology\cite{fontana}. One possibility to model proteins interactions is to use \emph{rules} to directly describe the evolution of a system. The next step is to see a \emph{mixture}, a group of proteins interacting with each others, as a graph: a node correspond to a protein with sites that can be linked and an edge to an interaction between sites of proteins. Finally, the evolution of a mixture can simply be described as a set of graph rewriting rules.

\Kappa{}\cite{kappa} is a graph rewriting language designed for this approach. It comes with an efficient simulator, which makes it a good basis for a formalism of molecular biology knowledge.

\subsection{Toward a first-order logic}
 Sadly, two problems arise when working with rule-bases models:
\begin{itemize}
\item Systems quickly become too large to be understood.
\item Adding a new knowledge to an existing set of rule is hard, since it will often require to change existing rules.
\end{itemize}

Husson worked in his thesis\cite{husson} to provide a logical framework that addresses these
issues by being expressive enough to represent molecular interactions while being simple enough to be ``compiled'' to a set of graph rewriting rules. His logic is a kind of a first-order logic that can be used to give information about the current micture and its next state. This is achieved by a $\star$ symbol that can be applied to predicates. A stared predicate gives information about the \emph{next} state of the mixture, while an un-stared symbol gives information about the actual state of the mixture.\\

For example in this logic, one can say ``The site s of the protein x turns ON'' with the following formula:
$$ OFF(x.s) \land ON^\star(x.s) $$
This formula uses two predicates $OFF$ and $ON$. Notice the $\star$ symbol after the $ON$ predicate meaning that $x$ will be $ON$ in the \emph{next} mixture, while we require it to be $OFF$ in the actual mixture since OFF is not starred.\\

Moreover, Husson's logic is a first-order logic, meaning that one can quantifies over proteins. For example, to say that ``A kinase site s can only turn ON when its site t is linked to another protein'', one can write to following formula:
$$ \forall x\ (K(x) \land OFF(x.s) \land ON^\star(x.s)) \implies \exists y\ Link(x.t,y) $$
It means that for all $x$, if $x$ is a kinase (assertion verified by the predicate $K$), is actually OFF and will become ON in the next state, then it exists a protein $y$ such that the site $t$ of $x$ is linked to $y$.\\

Husson bundled some predicates in his logic, namely:
\begin{itemize}
\item $=$, the binary predicate for equality.
\item $Link$, a binary predicate to express that two sites are linked.
\item $Has$, a unary predicate to express the presence of a protein in the mixture.
\end{itemize}

Other \emph{unary} predicates can be added when required. Moreover, the logician can declare predicates as static to forbid the use of the $\star$ operator, or dynamic to allow its use.

\subsection{A safe subset}
Husson identified a ``safe'' subset of his logic for which we know an algorithm able to synthesize equivalent \Kappa{} rules. The syntax of this subset is the following\cite{husson}:
\label{hussonsyntax}

$$ \theta_1,... ,\theta_k \text{ ensure } I_1,...,I_k \text{ maintain } I_{k+1},...,I_{k'} $$
where:
$$ \theta ::= L | \forall x\ \alpha(x,y) \Rightarrow \theta | \exists x\ \alpha(x,y) \land \theta | \theta \land \theta | \theta \lor \theta \quad \quad \forall L \in \text{Lit},x,y,\alpha \in \text{Guard}$$

A guard is a binary predicate involving exactly its two arguments (here, there is only two possible predicates: equality and linkability). Note that since every quantified variable is guarded, this fragment is strictly less expressive than the full logic.

The ``independent'' and ``maintain'' parts are optional and made of formulas of the following form:
$$ I ::= \alpha_1(x,y_1) \Rightarrow ...\Rightarrow \alpha_q(x,y_q) \Rightarrow \theta $$
\label{criterion}
where $Vars(\theta) \subseteq \{x,y_1,...,y_q\}$ and for each dynamic literal $L \in \theta$ we have $Var(L) = \{x\}$.

\subsection{A "logical" programming language}
Unfortunately, it is tedious to write directly formulas in this subset:

\begin{enumerate}
\item The syntax of the fragment is pretty restricted.
\item Some criterion of validity are not syntactic and needs to be manually checked.
\item Hand-made formulas can produce an enormous amount of rules frequently equivalent to a smaller set.
\item One could want to add a syntax by comprehension rather than explicitly write everything.
\end{enumerate}

Hence, there was a need for an intermediate language that is still made for experts but allowing the programmer to focus on describing its system rather than dealing with the formalism of the logic.\\
\Iota{} is an experiment trying to address these difficulties while compiling to Husson's fragment:
\begin{enumerate}
\item The syntax is of the original fragment is extended to include negation, let-definitions, and more.
\item Criterion of validity are automatically checked and the language features a powerful type system to ensure that formulas make sense.
\item The compiler is able to make some optimization based on the type system.
\item The compiler comes with a guard inference algorithm to help the programmer to focus on its system rather than finding the right expression.
\end{enumerate}

The next parts of this manual are organised as follows: first we introduce and explain the configuration file, next we describe precisely the language, and finally, we explain how the compiler was implemented.

\section{Semantics of the configuration file}
A \Iota{} program needs to be configured. For this purpose, the user needs to provide a configuration file separated in four parts:
\begin{enumerate}
\item The definition of constants.
\item The definition of types of objects.
\item The definition of predicates about these types.
\item Constraints of the type system.
\end{enumerate}

The full syntax is given in Section \ref{subsec:syntaxconfig}.
\subsection{The maxdepth parameter}
\Iota{} allows the programmer to describe trees (a protein with two sites is a node with two leaves). These trees have \emph{bounded} size, thus one must specify the maximum size of a tree. This is used in particular when compiling the \ocaml{TLink} predicate, see Section \ref{para:tlink}.

\subsection{Types}
When its come to programming, significant tools are \emph{types}. \Iota{} use them extensively to ensure that formulas are well formed and to perform several optimizations.
Nodes, leaves and some operations on trees are thus \emph{typed}. These types are all defined in the configuration file, and are grouped in two families:
\begin{itemize}
\item Types without accessors; they will represent leaves (which corresponds to sites for proteins).
\item Types with accessors; they will represent nodes. An accessor indicates the type of children and how to access it. There is two type of accessors:
  \begin{itemize}
  \item A simple accessor composed by a name and a type.
  \item Multiple accessors, composed by a name, a number and a type. The syntax \ocaml{10 f to site} will allow the use, on a variable \ocaml{x}, of \ocaml{x.f:0}, \ocaml{x.f:1}, to \ocaml{x.f:9}. This is semantically equivalent to define 10 different accessors.
  \end{itemize}
\end{itemize}

\subsection{Predicates}
\label{subsec:predicates}
Predicates can be used to extract information about nodes and leaves. Some predicates are bundled into the \Iota{} syntax itself:
\begin{itemize}
\item \ocaml{Eq} is a binary predicate expressing the fact that two literals are equals. It can be applied to literals of any types.
\item \ocaml{Link} is a binary predicate expressing the fact that two literals are linked. It can be applied to literals of any types, except when its use was restricted in the configuration file.
\item \ocaml{Has} is a unary predicate expressing the \emph{presence} of a literal. It can be applied to literal of any types.
\item \ocaml{TLink<t1,t2>} is a binary predicate expressing the fact that the two trees of the literals are linked. Its arguments must be of type \ocaml{t1} and \ocaml{t2}. This predicate was not originally bundled in Husson's safe fragment, but can be compiled to it (see Section \ref{para:tlink}).
\end{itemize}

The user can specify other predicates using the configuration file. Note that user-defined predicates are \emph{typed}, meaning that, once defined, they can be applied only to one type of literal.

\paragraph{Static vs Dynamic}
Predicates are separated into two categories: static and dynamic. Dynamic predicates can be prepended with the symbol \ocaml{+} in the language (which corresponds to the $\star$ symbol of Husson). \ocaml{Eq}, \ocaml{Link}, \ocaml{TLink} and \ocaml{Has} are dynamic.

\subsection{Constraints}
The \ocaml{Link} predicate is fully polymorphic but one can constrain its use, and force some types to be linked only to some others. This is very useful to force a protein to be linked only to some other kind of protein.

\subsection{Example}
Here is a valid configuration file:
\begin{minted}[frame=lines]{ocaml}
let maxdepth = 3

type site
type rsite
type mol = m to mol | f to site | g to site
type bigmol = 5 tm to mol | rs to rsite

static E about mol
static S about site
static R about rsite
dynamic Active about mol
dynamic Open about site

Link rsite to site
\end{minted}
The maximum depth is at 3, meaning that all considered trees will only have branches with less than 3 nodes.\\

This file declares four types:

\begin{itemize}
\item A leaf \ocaml{site}.
\item A leaf \ocaml{rsite}.
\item A node \ocaml{mol} with three possible children:
  \begin{itemize}
  \item An other node \ocaml{mol}.
  \item Two sites \ocaml{f} and \ocaml{g}.
  \end{itemize}
\item A node \ocaml{bigmol} with five possible children of type \ocaml{mol}, and one of type \ocaml{rsite}
\end{itemize}

The last line specify a constraint: a leaf of type \ocaml{rsite} can only be linked to leaves of type \ocaml{site}.

\section{Semantics of a \Iota{} program}
Once the configuration is written, one can write a program using it. The syntax is fully described in Section \ref{subsec:syntaxprog}. A \Iota{} program is a file composed of four parts (the last three are exactly from Husson's thesis) :

\begin{enumerate}
\item A list of let-definitions.
\item A list of formulas (which may use the above definitions).
\item (optional) A list of formulas (which may use the above definitions) that must be ensured.
\item (optional) A list of formulas (which may use the above definitions) that must be maintained.
\end{enumerate}

\subsection{Propositional formulas}
\paragraph{Literals}
Formulas are about literals. A \Iota{} literal is either:
\begin{multicols}{2}
  \begin{itemize}
  \item A name.
  \item A child of a literal.
  \item The parent of a literal.
  \end{itemize}
\end{multicols}

\paragraph{Formulas}
Propositional formulas can be built over (fully applied) predicates on literals using usual combiners: \ocaml{not}, \ocaml{&&} and \ocaml{||}.

\paragraph{About the negation}
Husson's original syntax didn't included the negation (see Section \ref{hussonsyntax}). Fortunately, it can be added since every formula can be put in normal form (where the negation is only present on predicates) \emph{while} still being in Husson's fragment.

\paragraph{Examples}
Here are some examples of formulas using the previous configuration:
\begin{minted}[frame=lines]{ocaml}
Active(x);
Active(x) && not (Active (x.m));
Has(x) && (not (Open(x.t)) || +Link(x.t,y.g))
\end{minted}

\subsection{Toward first-order}
Then, one can construct first-order formulas using the two following structures:
\begin{itemize}
\item \ocaml{forall x guards(x,y) -> formula(x)}
\item \ocaml{exists x guards(x,y) && formula(x)}
\end{itemize}
Note that each quantified variable is \emph{guarded} by a (disjunctive) non-empty list of guards.\\
This syntax and requirement are directly inspired by Husson's ``safe'' fragment. The only difference is that the original syntax was designed with only one guard for each variable. Thankfully, the two are equivalent since for each predicate $A$, $B$ and $C$ over the variable $x$, we have the following logical equivalences:

\begin{align*}
  (\forall x, (A x \lor B x) \implies C x) &\iff (\forall x, A x \implies C x) \land (\forall x, B x \implies C x)\\
  (\exists x, (A x \lor B x ) \land C x) &\iff (\exists x, A x \land B x) \lor (\exists x, B x \land C x)
\end{align*}

Hence our syntax is not more general but is simpler to use. This also considerably helps when compiling the \ocaml{TLink} predicate (Section \ref{para:tlink}) and when we run our guard inference algorithm (Section \ref{subsec:guardinfer}).

\paragraph{Guard}
A guard is a binary predicate involving exactly its two arguments. In \Iota{}, a guard is either \ocaml{Eq}, \ocaml{Link} or \ocaml{TLink}.

\subsection{Writing programs}
Now, we can wire up all these definitions using a standard let-syntax. Let-definitions are of the following form:
\begin{minted}[frame=lines]{ocaml}
let e x = E(x)
let f h x y = (h x) && (e y)
\end{minted}
There is a few things to note:
\begin{itemize}
\item Definitions are not recursive and are parsed up-to-bottom. Thus, in the example, \ocaml{f} can use \ocaml{e} but the converse is false.
\item All free variables of right hand-side must be in the left hand-side, with some exceptions (see below).
\item The language is \emph{high-order}, meaning that one definition can take another definition as an argument.
\end{itemize}

\subsubsection{A relaxed rule for free variables}
When writing a definition, it is convenient to see free variables as \emph{existentially} quantified ones. The problem is that, in \Iota{}, every quantified variable must be guarded; thus we need to infer a guard for free variables in let-definitions. We give an algorithm in Section \ref{subsec:guardinfer} allowing one to write a definition such as:
\begin{minted}{ocaml}
let f x = Eq(x,z) && Link(x,z)
\end{minted}
which will be compiled to:
\begin{minted}{ocaml}
let f x = exists z Eq(x,z) && Link(x,z)
\end{minted}

\subsection{General parts}
A program is optionally followed by two lists:
\begin{enumerate}
\item A list of formulas that must be \emph{ensured}.
\item A list of formulas that must be \emph{maintained}.
\end{enumerate}

These lists are composed of formulas are universally quantified and must respect Husson's criterion (see Section \ref{subsec:verification}).

\subsection{Examples}
\label{configex}
\begin{minted}[frame=lines]{ocaml}
let activate_and_link x y = Link (x.f, y.g) && +Active(x)

in

Has(a) && (forall u Link(a,u) -> (E(u) || TLink<mol,mol>(x,u)));
(activate_and_link x y) || (activate_and_link x z)

ensure

Link (z.f, w.g) => (E(z) -> (E(w) && Open (z.f)))
\end{minted}


\section{The compilation back-end}
The compilation is divided into 5 steps:
\begin{enumerate}
\item Parsing.
\item Typechecking.
\item Compilation itself.
\item Guard inference.
\item Checking of the resulting structure.
\end{enumerate}

\subsection{Parsing}
The described syntax is parsed using the great \Menhir{}\cite{menhir}. A first pass is made after the parsing itself to ensure that the user uses the symbol \ocaml{+} only in front of declared dynamic predicates.

Concerning the syntax itself, an interesting feature is that Husson's language is strictly included in \Iota{}. This means that the target language of our compiler is a strict subset of its source language. Hence, efforts were made to make the compiler \emph{reentrant}, meaning that if a program $P$ compiles to a program $P'$, then $P'$ compiles (exactly) to itself.

\subsection{Typechecking}
To ensure the validity of programs, they are typechecked in a Hindley-Milner system\cite{hindley} using a modified algorithm W\cite{milner}.
This standard choice gives a well-studied and powerful type system with an efficient algorithm. Moreover, since the \Iota{} syntax is very strict, most of the inference is trivial.

There is a challenge to infer the type of an expression using \ocaml{parent} since its usage introduce requirements about its argument (in the expression \ocaml{parent(x)}, \ocaml{x} must not be a root) which is hard to check. This is bypassed using explicit typing of this operator.

\subsubsection{About constraints}
As mentioned before, the use of the \ocaml{Link} predicate can be constrained. The validity check is made by modifying the classical algorithm W implementation: let-definitions now carry a map to remember which variable is linked to which type. Moreover, the algorithm now tries at the end to instantiate each type variable (hence, if a program typecheck, we know fore sure that it exists at least a type for each variable respecting the links constraints).\\

In details, defined constraints give a function $type \to type\ list\ option$ which represents the ``linkability'' of a type:
\begin{itemize}
\item $None$ means that a node of this type can be linked to any node of any other type.
\item $Some\ xs$ means that a node of this type can only be linked to node that have its type in $xs$.
\end{itemize}
When typing with algorithm W, each definition carries a map of type $typevariable \to (typevariable, type)\ either$ registering which variable is linked to which type. Then, when discovering the ground type (a type without an arrow that is not a variable) corresponding to a type variable, we can ensure that constraints are respected.

For example (using the configuration given in Section \ref{configex}):
\begin{minted}[frame=lines]{ocaml}
let f x y z = R(x) && Link(x,y) && Link(y,z)
in
S(b) && f a b c
\end{minted}
The typing of \ocaml{f} generates (using Greek letters for type variable):\\

\begin{figure}[H]
  \centering
  \subfloat[The environment]{
    \begin{tabular}{l|l}
      Variable& Type\\
      \hline
      x&$rsite$\\
      y&$\alpha$\\
      z&$\beta$
    \end{tabular}
  }
  \quad\quad
  \subfloat[The link map]{
    \begin{tabular}{l|l}
      Type Variable&Links\\
      \hline
      $\alpha$ & $rsite, \beta$\\
      $\beta$ & $\alpha$
    \end{tabular}
  }
\end{figure}

Next, the typing of the main part gives:
\begin{figure}[H]
  \centering
  \subfloat[The configuration]{
    \begin{tabular}{l|l}
      Type& Type\\
      \hline
      $rsite$&$site$
    \end{tabular}
  }
  \quad\quad
  \subfloat[The environment]{
    \begin{tabular}{l|l}
      Variable& Type\\
      \hline
      a&$rsite$\\
      b&$site$\\
      c&$\beta$
    \end{tabular}
  }
  \quad\quad
  \subfloat[The link map]{
    \begin{tabular}{l|l}
      Type Variable&Links\\
      \hline
      $\beta$ & $site$
    \end{tabular}
  }
\end{figure}

Note that the line about the type variable $\alpha$ in tabular $(e)$ was dropped since we identified that $\alpha$ was instantiated here to $site$ and \emph{this specialization was legal regarding the configuration and the link map}.

We can then instantiate $\beta$ to any type, since the linkabilty of $site$ is not constrained.

\subsubsection{Handling free variables}
Unusually, the typechecking actually transform the inputed program. Indeed, free variables of definitions are handled specifically since we want to try to quantify them existentially. Sadly, the language is still very rich and doing the guard inference at this stage of the compilation will be difficult. Thus, we introduce a new quantifier which mark the presence of an unbounded variable for later inference.

Since all definitions will also be inlined later, we rename unbounded free variable with new names to ensure that there will be no name capture.

\subsection{Compilation}
The compilation itself is divided into sub-steps.

\paragraph{Inlining}
All the definitions of \Iota{} are \emph{pure}, ie. without any effects, and not recursive, so one can always replace a call to a function by its body with proper substitution. This is the first thing the compiler do, since it allows to totally remove let-definitions. There are two sides effects:

\begin{itemize}
\item Unused definitions are not included in the result of the compilation. They still need to typecheck.
\item The size of the produced formula can be large since many expressions will be duplicated.
\end{itemize}

\paragraph{Compilation of TLink}
\label{para:tlink}
The \ocaml{TLink} predicate is not included in the target language, so we need to compile it.
Saying that two trees are linked is equivalent to say that there is a node in the first tree linked to another node in the other tree. Hence the compilation is made by a disjunction of all possible couples.
Note that thanks to the type system, the compilation of \ocaml{TLink} produces only valid (with respect to the type system and user-defined constraints) possibilities.

For example, using the following configuration:
\begin{minted}[frame=lines]{ocaml}
let maxdepth = 2
type site
type mol = m to mol | f to site | g to site
Link site to site
\end{minted}

The expression \ocaml{TLink<mol,mol>(x,y)} will generate the following disjunction:
\begin{minted}{ocaml}
Link(x.m.g,y) || Link(x.m.f,y) || Link(x.g,y) || Link(x.f,y)
\end{minted}

Note that there is for example no \ocaml{Link(x,y)} since it would violate the constraint that a site can only be linked to another site.

\paragraph{Optimization}
The compiler can do some optimization. It actually does only one, by replacing expressions of the form
\ocaml{parent<t1,t2>(x.f)} by \ocaml{x}.
Note that this optimization is legal in the sense that, since the program typechecks, we know that \ocaml{x.f} is a valid object.

\subsection{Guard inference}
\label{subsec:guardinfer}

\paragraph{Normal form of formulas}
Any formula can be put in a \emph{normal form}, where negation is only present on leaves. First of all, we put each formula into normal form, to fall back into Husson's syntax and simplify the guard inference.

\paragraph{The guard inference algorithm}
We give a general recursive algorithm to infer a list of disjunctive guards for a variable in Husson's fragment.\\
We write $X,\tau \vdash_x G$ if $x$ is guarded in $\tau$ by the disjunction of $g\in G$ and for each $g\in G$, the second variable is not in the forbidden set $X$ nor was quantified in the formula.

\[\inference{(y=x \land z \notin X) \lor (z=x \land y \notin X)}{X, Link(y,z) \vdash_x \{ Link(y,z) \}}  \quad \inference{(y=x \land z \notin X) \lor (z=x \land y \notin X)}{X, TLink(y,z) \vdash_x \{ TLink(y,z)\}}\]
\[\inference{(y=x \land z \notin X) \lor (z=x \land y \notin X)}{X, Eq(y,z) \vdash_x \{ Eq(y,z) \}}\]
\[\inference{X,A \vdash_x G }{X, A\land B \vdash_x G} \quad \inference{X,B \vdash_x G }{X, A\land B \vdash_x G}\]
\[\inference{X,A \vdash_x G_1  \quad X,B \vdash_x G_2}
  {X, A\lor B \vdash_x G_1 \cup G_2}\]
\[\inference{X\cup \{y\},A \vdash_x G }{X, \forall y A \vdash_x G} \quad
\inference{X\cup \{y\},A \vdash_x G }{X, \exists y A \vdash_x G}\]

The set of forbidden variables is used to ensure that $x$ is guarded only by variables bound before the apparition of $x$ (because for each quantification, we add the quantified variable to the forbidden set).

\paragraph{How to use it}
When encountering an expression $\tau$ with free variables $x_1,...,x_n$, the compiler generates all possible permutations of $\llbracket 1,n \rrbracket$, and tries to find a permutation that can be used to find guards for every variable.\\
More precisely, for each permutation $\pi$, it runs the mentioned algorithm for each $x_i$ with a starting set of forbidden variables $X_i = \{x_{\pi(j)} | i<j\le n\}$. Indeed, the permutation will give the order of quantification, and we only want guards for a said variable $x$ with variables bounded \emph{before} $x$.

\subsection{Checking}
\label{subsec:verification}
The resulting structure must satisfy some properties that cannot be hard-coded in the syntax. We must ensure that every guard is indeed a well-formed guard, and that the two parts ``ensure'' and ``maintain'' are well-formed with respect to Husson's criterion (see Section \ref{criterion}). This pass only checks if these properties are respected.

\section{About the compiler}
\subsection{Compiling}
The compiler is written using \Ocaml{}\cite{ocaml}. It was designed to be built with \Dune{}\cite{dune}. It is itself separated into two components:

\begin{enumerate}
\item A library, located in \verb|lib/| were all the work is made. Its entry-point is \verb|lib/main.ml|. It is fully documented using \Odoc{}\cite{odoc}. Online documentation is available at: \url{https://nobrakal.github.io/iota/iota/}, or can be built using:
\begin{minted}{shell}
dune build @doc
\end{minted}
\item A front-end, located in \verb|bin/|, providing an executable able to launch the compilation of a file. It can be built using:
\begin{minted}{shell}
dune build bin/main.exe
\end{minted}
  The produced executable takes as argument the configuration file and the file itself. It outputs the result of the compilation or prints an error.
\end{enumerate}

\subsection{Testing}
Some efforts were made to be able to test the compiler. The test-suite is located in \verb|test/|. It can be ran using:
\begin{minted}{shell}
dune runtest
\end{minted}
For now, it ensures that files that must compile are indeed compiling, and files that must fail to compile are indeed failing. It also ensures that the compiler is \emph{reentrant}, i.e. that if a file compiles to a program $A$, then the compilation of $A$ gives the exact same program.

\section{Conclusion}
\Iota{} is a tiny but powerful language which make a step after Husson's one toward an accessible language for molecular biologists. It addresses a good number of practical issues with Husson's logic and, while still made for experts, gives the possibility to write big and correct formulas by delegating some tasks to the compiler.\\

Contrary to approaches like LittleB \cite{littleb} which is plain LISP, \Iota{} is a very small language. This smalness is also its strength: it is very simple to analyse it and compile it to another language, while analysing full LISP is awfully more complicated.\\

The compiler can still be improved in several ways:
\begin{itemize}
\item We think that heuristics can be made to generate formulas that will behave well with the compilation of the logic to actual \Kappa{} rules.
\item One might want to add explicit typing to the syntax of \Iota{}. For now, there is no simple way to force the type of a variable other than defining a dedicated predicate for each type. Adding a syntax for explicit typing is only a mechanical work.
\item Explicit typing was required when using the \ocaml{parent} operator. Maybe there is a way to relax this requirement (for example, when there is only one possible parent for this type) and let the inference engine do the work.
\item \Kappa{} can actually be used to model other systems than molecular biology. Following this interpretation, \Iota{} maybe applies well to these fields.
\end{itemize}

\section{Syntax of the language}
Nota Bene:
\begin{itemize}
\item \verb|[<rule>]| symbolise a list of \verb|<rule>|.
\item \verb|[<rule> // <sep>]| symbolise a list of \verb|<rule>| separated by \verb|<sep>|.
\end{itemize}

Comments à-la-OCaml can be used everywhere.

\subsection{Syntax of configuration file}
\label{subsec:syntaxconfig} Here is a BNF syntax for the configuration file.
\begin{grammar}
  <constants> ::=
  \lit{let} \lit{maxdepth} \lit{=} <int>

  <accessor> ::=
  <lident> \lit{to} <type>
  \alt <int> <lident> \lit{to} <type>

  <typedef> ::=
  \lit{type} <lident>
  \alt \lit{type} <lident> \lit{=} [<accessor> // \lit{|}]

  <preddef> ::=
  \lit{static} <hident> \lit{about} <lident>
  \alt \lit{dynamic} <hident> \lit{about} <lident>

  <constraint> ::=
  \lit{Link} <lident> \lit{to} [<lident> // \lit{|}]

  <config> ::=
  <constants> [<typedef>] [<preddef>] [<constraint>]
\end{grammar}

\subsection{Syntax of programs}
\label{subsec:syntaxprog} Here is a BNF syntax for a \Iota{} program.
\begin{grammar}
  <lident> ::= [a-z][a-zA-Z0-9]*

  <hident> ::= [A-Z][a-zA-Z0-9]*

  <type> ::= <lident>

  <term> ::= <lident>
  \alt `parent<' <type> `,' <type> `>(' <term> `)'
  \alt <term> `.' <lident> `:' <int>
  \alt <term> `.' <lident>

  <guard> ::= `Eq(' <term> `,' <term> `)'
  \alt `Link(' <term> `,' <term> `)'
  \alt `TLink<' <type> `,' <type> `>(' <term> `,' <term> `)'

  <predicate> ::= <guard>
  \alt `Has(' <term> `)'
  \alt <hident> `(' <term> `)'

  <formula> ::= <predicate>
  \alt `+'<predicate>
  \alt `not' <formula>
  \alt <formula> `&&' <formula>
  \alt <formula> `||' <formula>
  \alt `(' <formula> `)'

  <safe> ::= <formula>
  \alt <term>
  \alt <safe> `&&' <safe>
  \alt <safe> `||' <safe>
  \alt `forall' <lident> `(' [<guard> // \lit{||}] `)' `->' <safe>
  \alt `exists' <lident> `(' [<guard> // \lit{||}] `)' `&&' <safe>
  \alt <safe> <safe>
  \alt `(' <safe> `)'

  <letdef> ::= \lit{let} <lident> [<lident>] \lit{=} <safe>

  <general> ::=
  <guard> \lit{->} <general>
  \alt \lit{=>} <formula>

  <program> ::=
  [<letdef>] \lit{in} [<safe> // \lit{;}] \lit{ensure} [<general> // \lit{;}] \lit{maintain} [<general> // \lit{;}]

\end{grammar}

\bibliographystyle{plain}
\bibliography{publications}

\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
